---
title: "HAR Project"
author: "Enrique Sanqui"
date: "Wednesday, January 21, 2015"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
---

# Overview
This document outlines a methodology to use the Human Activity Recognition (HAR) data to predict one of six activities recorded by wearable accelerometers and shared in the "Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements" paper by Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. 


# Approach

## Data Preparation
We removed identifiers for the experiment record, which did not add value to the classification problem.

```{r echo=FALSE , message=FALSE}
library(caret)
library(nnet)
```

```{r echo=FALSE}
#setwd("./Project")
pml.dat <- read.csv("./Data/pml-training.csv")
(remove.columns <- names(pml.dat[1:7]))
```

In addition, features that are missing more than 19000 out of the 19622 measurements (as blanks or NA's) were removed, as variability was too low to add any value.

```{r}
# GET LIST OF COLUMNS IN pml-training.csv EXCLUDING COLUMNS WITH >19000 NA VALUES AND SPACES
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
pml.dat.summary <- data.frame(summary(pml.dat))
exclude.columns <- rbind( subset(pml.dat.summary , grepl("^:19([0-9]{3,})",trim(pml.dat.summary$Freq) ) , select="Var2") ,
                          subset(pml.dat.summary , grepl("^NA's   :19([0-9]{3,})",trim(pml.dat.summary$Freq) ) , select="Var2") )
(include.field <- names(pml.dat[,-c(exclude.columns$Var2,1:7)]))
```


## Partitioning the HAR Data
We partitioned the data into __training__ and __testing__ data frames.  The model was selected using the __training__ data and we used the __testing__ data to estimate the Out Of Sample (OOS) Error. 

```{r}
# PARTITION DATA INTO TRAINING and TESTING ####
set.seed(77777)
i.train.test <- createDataPartition(pml.dat$classe , p=0.8 , list=FALSE)
training <- pml.dat[i.train.test,include.field]
testing <- pml.dat[-i.train.test,include.field]
```


## Model Selection

To predict the correct activity we explored the use of different classification models with built in feature selections.  We then examined their accuracy and ROC.

* rt - 99.85% Accuracy with mtry=2 and 99.93% with mtry=27 for the OOS Error
* rpart - 50.38% Accuracy for the OOS Error
* Boruta - model ran too long
* AdaBoost - 28.26% Accuracy for the OOS Error

Out of the tested models the Random Forest (__rf__) model returned the highest accuracy.  
The __rf__ model was trained using a 10-fold cross-validation, which was repeated 10 times.  

```{r , eval=FALSE}
# FIT Random Forest ####
my.grid <- expand.grid(mtry = c(2,27,52) )

# http://topepo.github.io/caret/training.html#control
fitControl <- trainControl(method = "repeatedcv", 
                           number = 10,
                           repeats = 10,
                           verboseIter = T,
                           p = 0.7
)  

# Traning Random Forest
set.seed(77777)
rf.mod <- train(x=training[,-dim(training)[2]] , y=as.factor(training$classe) ,
                linout=FALSE , 
                method="rf" , 
                tuneGrid = my.grid ,
                trControl = fitControl 
)
```

```{r echo=FALSE , message=FALSE}
# LOAD THE SAVED MODEL
load("rf_mod_training_150120")
```

```{r echo=FALSE , fig.width=10 , fig.height=5}
rf.mod

trellis.par.set(caretTheme())
plot(rf.mod)
```


Due to the amount of time it takes to train the model, we also save the trained __rf__ model. We are able to load it later using the __load()__ command.
```{r eval=FALSE}
# SAVE rf.mod TRAINING OBJECT
save( rf.mod , file="rf_mod_training")
```






## Estimated Out Of Sample Error
To get an estimate of the Out Of Sample Error, we used the __testing__ data frame partitioned earlier.
```{r message=FALSE}
# RANDOM FOREST CONFUSION MATRIX
(confusionMatrix( predict(rf.mod, testing) , testing$classe ))
```


## Variable Importance
As outlined earlier, the model also has built-in feature selection, which we extracted and plotted below.
```{r fig.height=10 , fig.width=10 }
plot(varImp(rf.mod, scale = FALSE))
```

